---
title: "Lab 16 - Robust Statistics"
date: 2022/12/06
author: "Nonparametric statistics ay 2022/2023"
output:
  
  html_document: 
    df_print: paged
  pdf_document: default
  html_notebook: 
    df_print: paged
  word_document: default
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
library(rgl)
knitr::opts_chunk$set(echo = TRUE)
knitr::knit_hooks$set(webgl = hook_webgl)
ggplot2::theme_set(ggplot2::theme_bw())
```

```{css, echo=FALSE}
.extracode {
background-color: lightblue;
}
```

## Loading necessary libraries

```{r message=FALSE, warning=FALSE}
library(robustbase)
library(psych)
library(MASS)
library(ellipse)
library(here)
library(DescTools)
library(knitr)
library(RobStatTM)
```

## Robust univariate

### Explicit robust estimators

Following the definitions outlined during the very first class of the
module, explicit robust estimators are readily available in R. There is
not very much to say about it here, let us only try to recreate the
table presented in slide 18 of the *Introduction and explicit robust
estimators* section.

```{r}
x <-
  c(9.52, 9.68, 10.16, 9.96, 10.08, 9.99, 10.47, 9.91, 9.92, 15.21)

x_9 <- x[-10]

est_names <- c(
  "$\\bar{x}$ ",
  "$\\bar{x}_{0.1}$ ",
  "$\\tilde{x}_{0.1}$",
  "$Me(x)$",
  "$\\hat{\\sigma}$",
  "$MD(x)$",
  "$MAD(x)$",
  "$MADN(x)$",
  "$IQR(x)$"
)

explicit_est_x_9 <- c(mean(x_9),
    mean(x_9, trim = .1),
    psych::winsor.mean(x = x_9, trim = .1),
    median(x_9),
    sd(x_9),
    DescTools::MeanAD(x_9),
    mad(x_9,constant = 1),
    mad(x_9), # MADN
    IQR(x_9))

explicit_est_x <- c(mean(x),
  mean(x, trim = .1),
  psych::winsor.mean(x = x, trim = .1),
  median(x),
  sd(x),
  DescTools::MeanAD(x),
  mad(x,constant = 1),
  mad(x),
  IQR(x)
  )

kable(
  data.frame(est_names, explicit_est_x_9, explicit_est_x),
  digits = c(0, 3, 3),
  escape = FALSE
)
```

### Implicit robust estimators

Univariate M-estimators can efficiently be created by means of the
`robustbase` R package. It offers plenty of features (we will see some
more further on in the lab) to customize your M-estimator, by even
letting you define your own `psiFunc`, for constructing `psi_func`
objects. We are not dwelling into the details about it, but if you are
curious you can look up at the vignette available
[online](https://cran.r-project.org/web/packages/robustbase/vignettes/psi_functions.pdf).
Let us stick with the two we have encountered in class, namely Huber and
Tukey's. They can easily be plotted by means of:

```{r}
source(system.file("xtraR/plot-psiFun.R", package = "robustbase", mustWork=TRUE))
p.psiFun(seq(-3,3,length.out=100), "huber", par = 1.5, leg.loc="bottomright", main="T")
p.psiFun(seq(-5,5,length.out=100), "biweight", par = 1.5, leg.loc="bottomright",main="T")
```

We have already seen how to hard-code the iterative algorithm described
in slide 13, but with an additional (robust) scale estimate
$\hat{\sigma}$.

```{r}
manual_M_location <-
  function(x,
           k,
           type = c("Huber", "Tukey"),
           tol = sqrt(.Machine$double.eps),
           itermax = 1000) {
    
    
    mu <- mu_old <- median(x) # initial value
    rob_scale <- mad(x) # it will be kept fixed
    crit <- TRUE
    iter <- 0
    
    weigth_f <- switch (type,
                        "Huber" = function(x) pmin(1, k/abs(x)),
                        "Tukey" = function(x) ((1-(x/k)^2)^2)*(abs(x)<=k)
    )
    
    while(crit){
      w_i <- weigth_f((x-mu)/rob_scale)
      mu <- weighted.mean(x = x,w = w_i)
      err <- abs(mu-mu_old)
      
      mu_old <- mu
      iter <- iter+1
      
      crit <- (err > tol & iter < itermax)
    }
    list(mu=mu, s=rob_scale,it=iter)
  }
```

The `robustbase` package contains the `huberM` function for performing
M-Estimation of location with MAD scale by means of Huber function

```{r}
huberM(x = x,k = 1.5)
manual_M_location(x = x,k = 1.5,type = "Huber")
```

In order to employ other $\psi$-functions, as well as other estimator
types (S, MM, not covered in class), the `nlrob` is a very flexible
function for robust fitting. Honestly, I do not find it straightforward
to use, so we move directly to robust multivariate estimators.

## Robust multivariate

### A simple example

We consider a data frame with average brain and body weights for $62$
species of land mammals and three other animal types.

```{r}
data("Animals2")
# let us work with the natural logarithms of the weigths
Animals2$body <- log(Animals2$body)
Animals2$brain <- log(Animals2$brain)
plot(Animals2)
```

Does it look familiar? We have already seen a plot of these data during
the class. The aim is to derive robust location and scale estimators
employing the Minimum Covariance Determinant (MCD). There are several
ways of doing it in `R`, with different routines available. We make use
of the `robustbase` package, that contains the "Essential" Robust
Statistics tools allowing to analyze data with robust methods.

The needed function is `covMcd`, whose main arguments are:

-   `x`: a matrix of data points
-   `alpha`: numeric parameter controlling the size of the subsets over
    which the determinant is minimized; roughly `alpha*N`
-   `nsamp`: number of subsets used for initial estimates or a character
    vector specifying whether "best", "exact", or "deterministic"
    estimation should be performed

Clearly, the most important hyper-parameter to be selected is `alpha`,
as it approximately determines the size of the subsets used for
parameter estimation, and the consequent BP of the estimator. Let us set
`alpha=0.75`.

```{r}
fit_MCD <- covMcd(x = Animals2, alpha = .75, nsamp = "best")
fit_MCD
```

The `fit_MCD` object contains the raw and reweighted estimates of
location and scatter, the resulting robust Mahalanobis distances and the
best subset used for computing the raw estimates.

```{r}
ind_best_subset <- fit_MCD$best
N <- nrow(Animals2)
p <- ncol(Animals2)
plot(Animals2, col=ifelse(1:N%in%ind_best_subset,"black","red"),pch=19)
```

Notice that the raw MCD estimate of scatter is by default multiplied by
a consistency factor

$$
c(p, \alpha)=\frac{\alpha}{F_{\chi_{p+2}^{2}}\left(q_{p, \alpha}\right)}
$$ ($q_{p, \alpha}$ is the $\alpha$ level quantile of a $\chi^2_p$
distribution) and a finite sample correction factor, to make it
consistent at the normal model and unbiased at small samples.

```{r}
dplyr::near(fit_MCD$raw.center,colMeans(Animals2[ind_best_subset,]))
dplyr::near(fit_MCD$raw.cov,cov(Animals2[ind_best_subset,])*prod(fit_MCD$raw.cnp2))
h <- fit_MCD$quan
dplyr::near(fit_MCD$raw.cnp2[1],(h/N)/pchisq(qchisq(p = h/N,df = p),df = p+2))
```

Notice that by default, unless specifying`raw.only=TRUE`, `covMcd`
performs a reweighting step for improving the efficiency of the final
estimator. In details, the new weights are computed as follows:

$$
W\left(d^{2}\right)=I\left(d^{2} \leqslant \chi_{p, 0.975}^{2}\right)
$$ where $d^2$ is the squared Mahalanobis distance based on the scaled
raw MCD.

```{r}
ind_rew_obs <-
  which(
    mahalanobis(
      x = Animals2,
      center = fit_MCD$raw.center,
      cov = fit_MCD$raw.cov
    ) <= qchisq(p = .975, df = p)
  )
dplyr::near(fit_MCD$center,colMeans(Animals2[ind_rew_obs,]))
dplyr::near(fit_MCD$cov,cov(Animals2[ind_rew_obs,])*prod(fit_MCD$cnp2))
```

`robustbase` provides also a nice plotting method that lets us fully
explore the set of graphical tools we have seen in class in an almost
automatic way:

```{r}
plot(fit_MCD,classic=TRUE)
```

We see that the outliers do not influence the robust estimates: the same
can only be partially said when employing the classical estimators.

Everything could have as always been hard-coded:

```{r, class.source="extracode", eval=FALSE}
n <- nrow(Animals2)
sample_mean <- apply(Animals2, 2, mean)
sample_cov <- cov(Animals2)#*(n-1)/n # by default R computes the corrected sample variance

# MCD estimates
plot(Animals2, xlim=c(-10,15),ylim=c(-5,15))
lines(ellipse(x = fit_MCD$cov,centre=fit_MCD$center),lty=2, col="blue",type="l")
points(x=fit_MCD$center[1],y=fit_MCD$center[2], pch="x", col="blue", cex=2)

# ML estimates
lines(ellipse(x = sample_cov, centre=sample_mean),lty=2, col="red",type="l")
points(x=sample_mean[1],y=sample_mean[2], pch="x", col="red", cex=2)

# Robust estimates
plot(sqrt(fit_MCD$mah))
# Classical estimates
plot(sqrt(mahalanobis(x = Animals2,center = sample_mean, cov = sample_cov)))
```

### A somewhat less simple example

Let us now consider a dataset containing measurements of $p= 9$
characteristics of $n = 677$ diaphragm parts used in the production of
TV sets. Diaphragm are thin metal plates, molded by a press. The aim of
the multivariate analysis is to gain insight into the production process
and the interrelations between the nine measurements and to find out
whether deformations or abnormalities have occurred and why.

```{r}
data_philips <- readRDS(here::here("Block V - Robust statistics/data/data_philips.Rds"))
```

```{r}
pairs(data_philips)
```

There is no clear visible pattern in the data, and no outliers seem to
be present.

```{r}
fit_MCD <- covMcd(x = data_philips, alpha = .75)
fit_MCD
```

```{r}
plot(fit_MCD, classic=TRUE, labels.id=FALSE, which="distance")
plot(fit_MCD,labels.id=FALSE,which=c("dd"))
```

Robust distances report a strongly deviating group of outliers, ranging
from index $491$ to index $565$. The reason being that the process
changed after the first $100$ points, and between index $491–565$ it was
out of control.

## Robust regression

### A simple example

Let us consider the Hertzsprung-Russell Diagram Data already encountered
in the very first lab and consider the problem of regressing the
logarithm of light intensity as a function of the logarithm of the
effective temperature at the surface of the star.

We have already seen that the OLS estimator goes bananas

```{r}
plot(starsCYG)
fit_lm <- lm(log.light~log.Te, data=starsCYG)
abline(fit_lm, col="red", lwd=2)
text(starsCYG$log.Te, starsCYG$log.light, 1:nrow(starsCYG), pos=1)
```

Yet, if we were to look at standard diagnostic plots we would not grasp
how bad the situation actually is:

```{r}
plot(fit_lm)
```

We would however if we fitted the Least Median of Squares (LMS) and the
Least Trimmed Squares (LTS)

```{r}
fit_lms <- lmsreg(log.light~log.Te, data=starsCYG)

# This is better than the previous
fit_lts <- ltsReg(log.light~log.Te, alpha=.75,mcd=TRUE,data=starsCYG) #ltsreg in the MASS package uses an older (slower) implementation of the LTS estimator
```

```{r}
plot(starsCYG)
abline(fit_lm, col="red", lwd=2)
abline(fit_lms, col="darkblue", lwd=2)
abline(fit_lts, col="darkgreen", lwd=2)
legend("bottomleft", c('OLS', 'LMS', 'LTS'), lwd=rep(2,4), col=c("red", "darkblue", "darkgreen"))
```

As for the MCD estimator, `robustbase` provides plotting methods for LTS

```{r}
plot(fit_lts)
```

### A somewhat less simple example

We consider $40$ cases of a study on production waste and land use,
originally published in Golueke, C.G. and McGauhey, P.H. (1970),
Comprehensive Studies of Solid Waste Management. The response variable
is solid waste (millions of tons), while the remaining explanatory
variables are industrial land (acres), fabricated metals (acres),
trucking and wholesale trade (acres), retail trade (acres) and
restaurants and hotels (acres).

```{r}
data("waste")

fit_lts <- ltsReg(SolidWaste~., alpha=.75,mcd=TRUE,data=waste) 
summary(fit_lts)
```

```{r}
plot(fit_lts)
```
