---
title: "Nonparametric Statistics Labs. Assembled!"
output:
    html_document:
        toc: true
        toc_float: true
        number_sections: true
    pdf_document:
        toc: true
        toc_depth: 3
        number_section: true
---
<!-- Chunk options: https://rmarkdown.rstudio.com/lesson-3.html#:~:text=include%20%3D%20FALSE%20prevents%20code%20and,useful%20way%20to%20embed%20figures. -->

```{r setup, include=FALSE}
options(knitr.duplicate.label = "allow")
path_images = paste(here::here("images"),"/", sep="")
knitr::opts_chunk$set(fig.path = path_images,
                      echo = TRUE,
					  dev = "png",
					  dpi = 300,
					  cache = TRUE,
					  warning = FALSE,
					  message = FALSE)
library(rgl)
knitr::knit_hooks$set(webgl = hook_webgl)
```

# Depth measures

```{r child=here::here("Block I - Nonparametric data exploration","NPS-lab01_depth_measures.Rmd")}
```

# Depth measures (functional)

```{r child=here::here("Block I - Nonparametric data exploration","NPS-lab02_funct_depth_measures.Rmd")}
```

# Univariate tests

```{r child=here::here("Block II - Nonparametric Inference","NPS-lab03_univariate_tests.Rmd")}
```

# Multivariate tests

```{r child=here::here("Block II - Nonparametric Inference","NPS-lab04_multivariate_tests.Rmd")}
```

# Permutation tests (functional)

```{r child=here::here("Block II - Nonparametric Inference","NPS-lab_Permutation_Functional.Rmd")}
```

# ANOVA

```{r child=here::here("Block II - Nonparametric Inference","NPS-lab05_ANOVA.Rmd")}
```

# Permutational Testing in Regression

```{r child=here::here("Block II - Nonparametric Inference","NPS-lab06_Regression.Rmd")}
```

# Permutational Confidence Interval

```{r child=here::here("Block II - Nonparametric Inference","NPS-lab07_Confidence.Rmd")}
```

# Bootstrapping

```{r child=here::here("Block II - Nonparametric Inference","NPS-lab08_Bootstrap.Rmd")}
```

# Bootstrap Tests and p-values

```{r child=here::here("Block II - Nonparametric Inference","NPS-lab09_BootstrapCI.Rmd")}
```

# Polynomial and step regression

```{r child=here::here("Block III - Nonparametric regression","NPS-lab10_polynomial_step_regression.Rmd")}
```

# Local regression and splines

```{r child=here::here("Block III - Nonparametric regression","NPS-lab11_local_regression_splines.Rmd")}
```

# Generalized Additive Models (GAM)

```{r child=here::here("Block III - Nonparametric regression","NPS-lab12_gam.Rmd")}
```

# Survival analysis

```{r child=here::here("Block III - Nonparametric regression","NPS-lab13_surv_analysis.Rmd")}
```

# Scalar, Multivariate and Functional Conformal Prediction

```{r child=here::here("Block IV - Nonparametric Prediction","NPS-lab14_Conformal.Rmd")}
```

# Robust Statistics

```{r child=here::here("Block V - Robust statistics","NPS-lab16_Robust_statistics.Rmd")}
```

