---
title: "Lab 11 - Local regression and splines"
date: 2022/11/10
author: "Nonparametric statistics ay 2022/2023"
output:
  
  html_document: 
    df_print: paged
  pdf_document: default
  html_notebook: 
    df_print: paged
  word_document: default
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
library(rgl)
knitr::opts_chunk$set(echo = TRUE)
knitr::knit_hooks$set(webgl = hook_webgl)
ggplot2::theme_set(ggplot2::theme_bw())
```

```{css, echo=FALSE}
.extracode {
background-color: lightblue;
}
```

## Loading necessary libraries

```{r message=FALSE, warning=FALSE}
library(ISLR2)
library(car)
library(np)
library(splines)
library(fda)
library(magrittr)
library(KernSmooth)
```

## Nonparametric Regression (cont.)

During the last lab we have familiarized with the two simplest ways to
move beyond linearity in (univariate) regression, namely with polynomial
and step functions. In this lab, we will look at two more methods that
provide non-linear fitting: local regression and splines. While the
former relies on local fitting (as the name suggests), splines will
allow to merge together polynomial and step functions providing a
powerful linear smoother.

## Local regression

Local regression involves computing the fit at a target point $x_0$
using only the nearby training observations. Local regression is
sometimes referred to as a *memory-based* procedure because, like
nearest-neighbors, we need all the training data each time we wish to
compute a prediction. Let us keep working with the Prestige dataset. We
start with weighted local averaging with uniform (or rectangular)
kernel.

```{r message=FALSE, warning=FALSE}
m_loc = npreg(prestige ~ income,
              ckertype = 'uniform',
              bws = 3200, # bandwidth
              data = Prestige)

income_newdata=data.frame(income=with(Prestige, seq(range(income)[1],range(income)[2],by=0.5)))
preds=predict(m_loc,newdata=income_newdata,se=T)
se.bands=cbind(preds$fit +2* preds$se.fit ,preds$fit -2* preds$se.fit)
with(
  Prestige,
  plot(
    income ,
    prestige ,
    xlim = range(income_newdata$income) ,
    cex = .5,
    col = " darkgrey ",
    main = 'Local Averaging - bws3200 - Uniform kernel'
  )
)
lines(income_newdata$income,preds$fit ,lwd =2, col =" blue")
matlines(income_newdata$income,se.bands ,lwd =1, col =" blue",lty =3)
```

This can be manually implemented in a straightforward way. Spoiler
alert: the code below is very bad in terms of efficiency and
portability, it is only shown for didactic purposes!

```{r, class.source="extracode"}
loc_pred_unif_manual <- Vectorize(function(x_0, bw) {
  
  ind_unit_in_bw <-
    which(Prestige$income > (x_0 - bw) & Prestige$income < (x_0 + bw))
  
  Prestige %>%
    dplyr::slice(ind_unit_in_bw) %>%
    dplyr::summarise(mean(prestige)) %>%
    dplyr::pull()
  
}, vectorize.args = "x_0")


# Predict original data

preds <- predict(m_loc, newdata=data.frame(income=Prestige$income))

all(dplyr::near(preds, loc_pred_unif_manual(x_0 = Prestige$income,bw = 3200), tol = 1e-13))

```

Let us try to decrease the bandwidth

```{r message=FALSE, warning=FALSE}
m_loc = npreg(prestige ~ income,
              ckertype = 'uniform',
              bws = 1000, # bandwidth
              data = Prestige)

income_newdata=data.frame(income=with(Prestige, seq(range(income)[1],range(income)[2],by=0.5)))
preds=predict(m_loc,newdata=income_newdata,se=T)
se.bands=cbind(preds$fit +2* preds$se.fit ,preds$fit -2* preds$se.fit)
with(
  Prestige,
  plot(
    income ,
    prestige ,
    xlim = range(income_newdata$income) ,
    cex = .5,
    col = " darkgrey ",
    main = 'Local Averaging - bws1000 - Uniform kernel'
  )
)
lines(income_newdata$income,preds$fit ,lwd =2, col =" blue")
matlines(income_newdata$income,se.bands ,lwd =1, col =" blue",lty =3)
```

We have issues with uniform kernel if there are no data in the bins...

```{r message=FALSE, warning=FALSE}
m_loc = npreg(prestige ~ income,
              ckertype = 'uniform',
              bws = 5000, # bandwidth
              data = Prestige)

income_newdata=data.frame(income=with(Prestige, seq(range(income)[1],range(income)[2],by=0.5)))
preds=predict(m_loc,newdata=income_newdata,se=T)
se.bands=cbind(preds$fit +2* preds$se.fit ,preds$fit -2* preds$se.fit)
with(
  Prestige,
  plot(
    income ,
    prestige ,
    xlim = range(income_newdata$income) ,
    cex = .5,
    col = " darkgrey ",
    main = 'Local Averaging - bws5000 - Uniform kernel'
  )
)
lines(income_newdata$income,preds$fit ,lwd =2, col =" blue")
matlines(income_newdata$income,se.bands ,lwd =1, col =" blue",lty =3)
```

An option could be to change kernel:

```{r message=FALSE, warning=FALSE}
m_loc = npreg(prestige ~ income,
              ckertype = 'gaussian',
              bws = 3200, # bandwidth
              data = Prestige)

income_newdata=data.frame(income=with(Prestige, seq(range(income)[1],range(income)[2],by=0.5)))
preds=predict(m_loc,newdata=income_newdata,se=T)
se.bands=cbind(preds$fit +2* preds$se.fit ,preds$fit -2* preds$se.fit)
with(
  Prestige,
  plot(
    income ,
    prestige ,
    xlim = range(income_newdata$income) ,
    cex = .5,
    col = " darkgrey ",
    main = 'Local Averaging - bws3200 - Gaussian kernel'
  )
)
lines(income_newdata$income,preds$fit ,lwd =2, col =" blue")
matlines(income_newdata$income,se.bands ,lwd =1, col =" blue",lty =3)
```

Several R packages provide implementations of local polynomial
estimators. For instance, notice that the same results could be achieved
by means of the `locpoly` function in the `KernSmooth` package:

```{r}
m_kern_smooth <- with(Prestige,KernSmooth::locpoly(x = income, y = prestige, bandwidth = 3200, degree = 0,
                           range.x = c(range(income)[1], range(income)[2]), gridsize = nrow(income_newdata)))

all(dplyr::near(m_kern_smooth$y,preds$fit,tol = 1e-1)) # somewhat different approx up to .1
```

Furthermore, also the built-in `loess` function can be used for local
polynomial regression fitting. Contrarily to `locpoly` and `npreg`,
`loess` has a different control of the bandwidth by means of the span
argument, defining the proportion of points of the sample that are taken
into account for performing the local fit. Additionally, `loess` uses a
triweight kernel for weighting the points contributions.

```{r message=FALSE, warning=FALSE}
m_loess = loess(prestige ~ income,
                span = .35, degree=0,
              data = Prestige)
preds_loess=predict(m_loess,newdata=income_newdata,se=T)
se.bands_loess=cbind(preds_loess$fit +2* preds_loess$se.fit ,preds_loess$fit -2* preds_loess$se.fit)
with(
  Prestige,
  plot(
    income ,
    prestige ,
    xlim = range(income_newdata$income) ,
    cex = .5,
    col = " darkgrey ",
    main = 'Local Polynomial Regression - span= 0.35 - Triweight kernel'
  )
)
lines(income_newdata$income,preds_loess$fit ,lwd =2, col =" green")
matlines(income_newdata$income,se.bands_loess ,lwd =1, col =" green",lty =3)
```

Last note: both `locpoly` and `loess` have a degree argument that allows
us to control the degree of local polynomial used in the estimation. In
the previous examples we have always relied on local constant fitting
setting `degree=0` (also known as the Nadaraya--Watson estimator), but
polynomial with higher degrees can be used. If you want to have a very
nice interactive visualization on how local regression works you can
check the following
[link](https://bookdown.org/egarpor/PM-UC3M/npreg-kre.html#npreg-kre-locpoly)
(Figure 6.6).

### Exercise

Try to perform again the analysis employing the last continuous kernel
type available in the `np` package, namely the Epanechnikov kernel. Do
the results change much with respect to the Gaussian kernel?

## Splines

Splines are piecewise polynomial functions which are constrained to join
smoothly at knots. Nevertheless, before looking at splines let us
empirically motivate their need by considering piecewise polynomials
first. Let us keep working with our Prestige dataset. Let us hard-code a
piecewise polynomial with discontinuity at the chosen cutoff (hereafter
called knot)

```{r}
cutoff <- 10000
Prestige$income_cut <- Prestige$income>cutoff

Prestige$income_cut_model <- (Prestige$income-cutoff)*Prestige$income_cut

model_cut_disc=lm(prestige ~ income + income_cut_model + I(income>cutoff), data=Prestige)
# model_cut_disc=lm(prestige ~ income * income_cut, data=Prestige) # equivalent formulation

new_data <-
  with(Prestige, data.frame(
    income = seq(range(income)[1], range(income)[2], by = 0.5)
  ))

new_data$income_cut_model = (new_data$income - cutoff) * (new_data$income > cutoff)

preds=predict(model_cut_disc,new_data ,se=T)
se.bands=cbind(preds$fit +2* preds$se.fit ,preds$fit -2* preds$se.fit)

with(Prestige, plot(income ,prestige ,xlim=range(new_data$income) ,cex =.5, col =" darkgrey " ))
lines(new_data$income,preds$fit ,lwd =2, col =" blue")
matlines(new_data$income, se.bands ,lwd =1, col =" blue",lty =3)
```

We can easily consider polynomials that go beyond the first degree:

```{r}

model_cut_quad <- lm(prestige ~ poly(income,degree = 2) + income_cut_model + I(income>cutoff),  data=Prestige)
preds=predict(model_cut_quad,new_data,se=T)
se.bands=cbind(preds$fit +2* preds$se.fit ,preds$fit -2* preds$se.fit)

with(Prestige, plot(income ,prestige ,xlim=range(new_data$income) ,cex =.5, col =" darkgrey " ))
lines(new_data$income,preds$fit ,lwd =2, col =" blue")
matlines(new_data$income, se.bands ,lwd =1, col =" blue",lty =3)
```

This indeed gives us a lot of flexibility, yet you can feel that having
discontinuities in the estimated regression line may not be ideal... We
can force the regression line to be continuous by dropping the
`I(income>cutoff)` term in the `model_cut_disc` model, thus getting back
a degree of freedom:

```{r}
model_cut=lm(prestige ~ income + income_cut_model, data=Prestige)
preds_cut=predict(model_cut,new_data ,se=T)
se.bands_cut=cbind(preds_cut$fit +2* preds_cut$se.fit ,preds_cut$fit -2* preds_cut$se.fit)
with(Prestige, plot(income ,prestige ,xlim=range(new_data$income) ,cex =.5, col =" darkgrey " ))
lines(new_data$income,preds_cut$fit ,lwd =2, col =" blue")
matlines(new_data$income, se.bands_cut ,lwd =1, col =" blue",lty =3)
```

We have manually fitted our first piece-wise linear spline with one knot
at $x=10000$! How to move forward in this direction? That is, allowing
for different functional forms in different bins whilst maintaining some
degree of smoothness in the interpolation? We can effectively do so by
employing the built-in `splines` package. The idea underlying regression
splines relies on specification of a set of knots, producing sequence of
basis functions and then using least squares for estimating
coefficients.

```{r}
model_linear_spline <- lm(prestige ~ bs(income, knots=10000,degree=1), data=Prestige)

preds=predict(model_linear_spline,new_data,se=T)
se.bands=cbind(preds$fit +2* preds$se.fit ,preds$fit -2* preds$se.fit)

with(Prestige, plot(income ,prestige ,xlim=range(new_data$income) ,cex =.5, col =" darkgrey " ))
lines(new_data$income,preds$fit ,lwd =2, col =" blue")
matlines(new_data$income, se.bands ,lwd =1, col =" blue",lty =3)
```

Recall that we can always look at the design matrix (and how the `bs()`
command builds the polynomial splines) with the `model.matrix()`
function

```{r}
head(unname(model.matrix(model_linear_spline)))
```

Notice that predictions made with the hard-coded `model_cut` model is
the same as the `model_linear_spline` one:

```{r}
all(dplyr::near(x = preds_cut$fit,preds$fit))
```

Yet, the model coefficients are different:

```{r}
rbind(model_cut=coef(model_cut),
model_linear_spline=coef(model_linear_spline))
```

The reason is due to the different set of basis used to represent the
space of spline functions: the former employs a conceptually simple (and
easy to code) truncated power basis, whereas the latter uses a
computationally more efficient B-spline basis (more on this later in the
lab).

Flexibility? The sky is the limit here. Let us build a piecewise linear
spline

```{r}
inc_breaks <- c(seq(0, 10000, by = 2500)[-1], 15000)

model_linear_spline_2 <-
  lm(prestige ~ bs(income, knots = inc_breaks, degree = 1), data = Prestige)

preds=predict(model_linear_spline_2, new_data,se=T)
se.bands=cbind(preds$fit +2* preds$se.fit ,preds$fit -2* preds$se.fit)

with(Prestige, plot(income ,prestige ,xlim=range(new_data$income) ,cex =.5, col =" darkgrey " ))
lines(new_data$income,preds$fit ,lwd =2, col =" blue")
matlines(new_data$income, se.bands ,lwd =1, col =" blue",lty =3)
```

Piecewise quadratic

```{r}
model_quad_splines <-
  lm(prestige ~ bs(income, knots = inc_breaks, degree = 2), data = Prestige)

preds=predict(model_quad_splines, new_data,se=T)
se.bands=cbind(preds$fit +2* preds$se.fit ,preds$fit -2* preds$se.fit)

with(Prestige, plot(income ,prestige ,xlim=range(new_data$income) ,cex =.5, col =" darkgrey " ))
lines(new_data$income,preds$fit ,lwd =2, col =" blue")
matlines(new_data$income, se.bands ,lwd =1, col =" blue",lty =3)
```

Cubic spline

```{r}
model_cubic_splines <-
  lm(prestige ~ bs(income, knots = inc_breaks, degree = 3), data = Prestige)

preds=predict(model_cubic_splines, new_data,se=T)
se.bands=cbind(preds$fit +2* preds$se.fit ,preds$fit -2* preds$se.fit)

with(Prestige, plot(income ,prestige ,xlim=range(new_data$income) ,cex =.5, col =" darkgrey " ))
lines(new_data$income,preds$fit ,lwd =2, col =" blue")
matlines(new_data$income, se.bands ,lwd =1, col =" blue",lty =3)
```

I can either specify the knots (as done so far), or have R automatically
select the spacing, I just need to specify their number knowing that:
$$dof = \#knots + degree$$

Example: if we want $4$ knots and a cubic spline, we will end up with
$7$ degrees of freedom.

```{r}
model_cubic_splines_2 <-
  lm(prestige ~ bs(income, degree = 3,df = 7), data = Prestige)

preds=predict(model_cubic_splines_2, new_data,se=T)
se.bands=cbind(preds$fit +2* preds$se.fit ,preds$fit -2* preds$se.fit)

with(Prestige, plot(income ,prestige ,xlim=range(new_data$income) ,cex =.5, col =" darkgrey " ))
lines(new_data$income,preds$fit ,lwd =2, col =" blue")
matlines(new_data$income, se.bands ,lwd =1, col =" blue",lty =3)

knots <- attr(bs(Prestige$income, degree=3,df=7),'knots')
knots_pred=predict(model_cubic_splines_2,list(income=knots))
points(knots,knots_pred, col='blue',pch=19)
abline(v = knots, lty=3)
```

We can visualize how B-spline bases look like in the unit interval:

```{r, fig.asp = 1.3}
splines_grid <- seq(0, 1, by=0.001)

lin_spline <- bs(splines_grid, df=7, degree = 1)
quad_spline <- bs(splines_grid, df=8, degree = 2)
cub_spline <- bs(splines_grid, df=9, degree = 3)

knots_splines <- attributes(lin_spline)$knots

par(mfrow=c(3,1))
plot(lin_spline[,1]~splines_grid, ylim=c(0,max(lin_spline)), type='l', lwd=2, col=1, 
     xlab="", ylab="", main="Linear B-spline basis")
for (j in 2:ncol(lin_spline)) lines(lin_spline[,j]~splines_grid, lwd=2, col=j)

points(knots_splines,rep(0,length(knots_splines)),pch=19)
abline(v = knots_splines,lty=3)

plot(quad_spline[,1]~splines_grid, ylim=c(0,max(quad_spline)), type='l', lwd=2, col=1, 
     xlab="", ylab="", main="Quadratic B-spline basis")
for (j in 2:ncol(quad_spline)) lines(quad_spline[,j]~splines_grid, lwd=2, col=j)
points(knots_splines,rep(0,length(knots_splines)),pch=19)
abline(v = knots_splines,lty=3)


plot(cub_spline[,1]~splines_grid, ylim=c(0,max(cub_spline)), type='l', lwd=2, col=1, 
     xlab="", ylab="", main="Cubic B-spline basis")
for (j in 2:ncol(cub_spline)) lines(cub_spline[,j]~splines_grid, lwd=2, col=j)
points(knots_splines,rep(0,length(knots_splines)),pch=19)
abline(v = knots_splines,lty=3)
```

```{r, echo=FALSE,eval=TRUE}
par(mfrow=c(1,1))
```

Notice that, by employing the user-friendly truncated power basis,

```{=tex}
\begin{aligned}
h_{j}(X) &=X^{j-1}, j=1, \ldots, M \\
h_{M+\ell}(X) &=\left(X-\xi_{\ell}\right)_{+}^{M-1}, \ell=1, \ldots, K
\end{aligned}
```
where M denotes the order of the spline, $\xi_{\ell}$ the $l$-th knot
and $\left(A\right)_{+}=\max(A, 0)$, we can actually hard code the same
model as `model_cubic_splines_2`

```{r, class.source="extracode"}
# Define trunc power basis
M <- 4 # cubic spline
X_powers <- poly(Prestige$income,degree = (M-1),raw = TRUE)
X_trunc_powers <- sapply(knots, function(k) (Prestige$income-k)^(M-1)*(Prestige$income>k))
X_manual_cubic_splines <- cbind(X_powers, X_trunc_powers)

# Fit the model with OLS
model_cubic_splines_2_manual <- lm(Prestige$prestige ~ X_manual_cubic_splines)

# Construct the new_data_manual object
new_data_manual_powers <- poly(new_data[,1],degree = 3,raw = TRUE)
new_data_manual_trunc_powers <- sapply(knots, function(k) (new_data[,1]-k)^3*(new_data[,1]>k))
new_data_manual <- cbind(new_data_manual_powers, new_data_manual_trunc_powers) 

# Make predictions

preds_manual_cubic_splines <- c(cbind(1,new_data_manual)%*%model_cubic_splines_2_manual$coefficients)

# Compare with the previous solution
all(dplyr::near(preds$fit,preds_manual_cubic_splines))
```

Apart from the considerable coding effort with respect to using the
built-in `bs()` function, values in the coefficients vector and in the
design matrix get absurdly low and high in our hard-coded solution...
Let us stick with B-spline bases from now on!

How to avoid weird behavior at the boundaries of the domain? We can
employ natural splines

```{r}
knots <- quantile(Prestige$income,probs=c(0.1,0.5,0.9))
boundary_knots <- quantile(Prestige$income,probs=c(0.05,0.95))

model_ns=lm(prestige ~ ns(income,knots=knots,Boundary.knots=boundary_knots), data=Prestige) #defaults to three knots
preds=predict(model_ns, new_data,se=T)
se.bands=cbind(preds$fit +2* preds$se.fit ,preds$fit -2* preds$se.fit)

with(Prestige, plot(income ,prestige ,xlim=range(new_data$income) ,cex =.5, col =" darkgrey " ))
lines(new_data$income,preds$fit ,lwd =2, col =" blue")
matlines(new_data$income, se.bands ,lwd =1, col =" blue",lty =3)

knots_pred=predict(model_ns,list(income=knots))
points(knots,knots_pred, col='blue',pch=19)
boundary_pred <- predict(model_ns,list(income=boundary_knots))
points(boundary_knots,boundary_pred,col='red',pch=19)
abline(v = knots, lty=3, col="blue")
abline(v = boundary_knots, lty=3, col="red")
```

Clearly, the specification of knots can be non-quantile driven in
general.

### Exercise

Play around with the splines functions, applying them to the Wage
dataset.

```{r}
data(Wage)
with(Wage, plot(age,wage))
```

## Smoothing splines

With smoothing splines we look for a function $f(\cdot)$ that makes
$RSS=\sum_{i=1}^n(y_i-f(x_i))^2$ small, but that is also smooth. That
is, we aim at minimizing

```{=tex}
\begin{equation}
\operatorname{RSS}(f, \lambda)=\sum_{i=1}^{n}\left\{y_{i}-f\left(x_{i}\right)\right\}^{2}+\lambda \int\left\{f^{\prime \prime}(t)\right\}^{2} d t
\end{equation}
```
This is operationally done by performing a penalized regression over the
natural spline basis, placing knots at all the inputs. This means that,
remarkably, the problem defined on an infinite-dimensional function
space has a finite-dimensional, unique minimizer which is a natural
cubic spline! This result is not even difficult to prove (see Theorem
2.3 of Nonparametric Regression and Generalized Linear Models: A
Roughness Penalty Approach by Green and Silverman) This is automatically
performed in R through the `smooth.spline` function.

```{r}
fit_smooth_spline <- with(Prestige, smooth.spline(income,prestige,df=100))
with(Prestige, plot(income ,prestige, cex =.5, col =" darkgrey "))
lines(fit_smooth_spline,col="blue",lwd=2)
```

```{r}
fit_smooth_spline <- with(Prestige, smooth.spline(income,prestige,df=20))
with(Prestige, plot(income ,prestige, cex =.5, col =" darkgrey "))
lines(fit_smooth_spline,col="blue",lwd=2)
```

Or we can directly specify the value of the smoothing parameter
$\lambda$

```{r}
fit_smooth_spline <- with(Prestige, smooth.spline(income,prestige,lambda = 1e-1))
with(Prestige, plot(income ,prestige, cex =.5, col =" darkgrey "))
lines(fit_smooth_spline,col="blue",lwd=2)
```

```{r}
fit_smooth_spline <- with(Prestige, smooth.spline(income,prestige,lambda = 1e-6))
with(Prestige, plot(income ,prestige, cex =.5, col =" darkgrey "))
lines(fit_smooth_spline,col="blue",lwd=2)
```

Generally, one wants to optimize the LOOCV error

```{=tex}
\begin{equation}
\mathcal{V}_{o}=\frac{1}{n} \sum_{i=1}^{n}\left(\hat{f}_{i}^{[-i]}-y_{i}\right)^{2}=\frac{1}{n} \sum_{i=1}^{n}\left(y_{i}-\hat{f}_{i}\right)^{2} /\left(1-A_{i i}\right)^{2}
\end{equation}
```
or the GCV error

```{=tex}
\begin{equation}
\mathcal{V}_{g}=\frac{n \sum_{i=1}^{n}\left(y_{i}-\hat{f}_{i}\right)^{2}}{[n-\operatorname{tr}(\mathbf{A})]^{2}}
\end{equation}
```
where $\boldsymbol{A}$ is the hat matrix of the associated regression
problem (see e.g., Section 5.4.1 of The Elements of Statistical Learning
by Hastie, Tibshirani and Friedman for further explanation).

```{r}
fit_smooth_spline_CV <- with(Prestige, smooth.spline(income,prestige,cv = TRUE))
fit_smooth_spline_GCV <- with(Prestige, smooth.spline(income,prestige,cv = FALSE))
with(Prestige, plot(income ,prestige, cex =.5, col =" darkgrey "))
lines(fit_smooth_spline_CV,col="red",lwd=2,lty=1)
lines(fit_smooth_spline_GCV,col="blue",lwd=2, lty=2)
legend(20000, 30, legend=c("CV", "GCV"),
       col=c("red", "blue"), lty=1:2, cex=0.8)
```

Likewise for the previous methods, it is very easy to produce forecasts

```{r}
predict(fit_smooth_spline_GCV,22000)
```
