---
title: "Lab 12 - Generalized Additive Models (GAM)"
date: 2022/11/11
author: "Nonparametric statistics ay 2022/2023"
output:
  
  html_document: 
    df_print: paged
  pdf_document: default
  html_notebook: 
    df_print: paged
  word_document: default
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
library(rgl)
knitr::opts_chunk$set(echo = TRUE)
knitr::knit_hooks$set(webgl = hook_webgl)
ggplot2::theme_set(ggplot2::theme_bw())
```

```{css, echo=FALSE}
.extracode {
background-color: lightblue;
}
```

## Loading necessary libraries

```{r message=FALSE, warning=FALSE}
library(ISLR2)
library(car)
library(mgcv)
library(rgl)
library(splines)
library(pbapply)
```

## Moving beyond simple (nonparametric) regression

In the past two labs we have seen several methods that allow us to fit a
smooth regression function
$$
y_i=f(x_i)+\varepsilon_i, \quad i=1,\ldots,N
$$
with $x_i\in \mathbb{R}$. In today's lab, we would like to move forward
looking at a way to perform multivariate nonparametric regression, that
is when $x_i\in \mathbb{R}^p$, $p>1$. We will do so by means of an
additive approach through the Generalized Additive Models. Let us go
back to our well-known Prestige dataset, but let us consider the
education variable as well.

```{r}
with(Prestige, scatterplotMatrix(data.frame(prestige, education, income)))
```

The relationship between prestige and education seems fairly linear. On
the contrary, as we have extensively experienced, prestige and income
are quite nonlinearly related.

First off, let's try to fit a multivariate linear model

```{r}
model_lm=lm(prestige ~ education + income, data=Prestige)
summary(model_lm)
```

Building prediction surfaces is as simple as it was for the
bidimensional case

```{r}
education.grid=seq(range(Prestige$education)[1],range(Prestige$education)[2],length.out = 100)
income.grid=seq(range(Prestige$income)[1],range(Prestige$income)[2],length.out = 100)
grid=expand.grid(education.grid,income.grid)
names(grid)=c('education','income')
pred=predict(model_lm,newdata=grid) 
```

```{r, webgl=TRUE}
persp3d(education.grid,income.grid,pred,col='blue',border="black",lwd=0.3)
with(Prestige, points3d(education,income,prestige,col='black',size=5))
```

What happens if I add the interaction?

```{r}
model_lm_interaction <- lm(prestige ~ education + income + education:income, data=Prestige) 
# model_lm_interaction <- lm(prestige ~ education * income, data=Prestige) # alternatively 
summary(model_lm_interaction)
```

I manage to capture a bit of the nonlinearity with the interaction
term... Can we do better?

```{r, webgl=TRUE}
pred_interaction=predict(model_lm_interaction,newdata=grid)
persp3d(education.grid,income.grid,pred_interaction,col='grey30')
with(Prestige,points3d(education,income,prestige,col='black',size=5))
```

We certainly can by employing Generalized Additive Models! The main
reference (also, surprisingly easy and straightforward to use) is
[Generalized Additive Models: An Introduction with
R](https://www.taylorfrancis.com/books/mono/10.1201/9781315370279/generalized-additive-models-simon-wood)
by Simon N. Wood. The package `mgcv` accompanies the book, it is a
very-well structured and actively maintained package that provides
computation for smoothness estimation for a variety of models: we will
only scratch the surface today! Recall the analytical expression of a
GAM:

```{=tex}
\begin{aligned}
y_{i} &=\beta_{0}+\sum_{j=1}^{p} f_{j}\left(x_{i j}\right)+\epsilon_{i} \\
&=\beta_{0}+f_{1}\left(x_{i 1}\right)+f_{2}\left(x_{i 2}\right)+\cdots+f_{p}\left(x_{i p}\right)+\epsilon_{i}
\end{aligned}
```
It is called an *additive* model because we calculate a separate $f_{j}$
for each $X_{j}$, and then add together all of their contributions.
Operatively, `gam()` is the function of the `mgcv`package for fitting
these types of models. It works very much like the `glm` function for
generalized linear models. Clearly with the former we are allowed to
specify diverse smooths options for the model terms.

```{r}
model_gam=gam(prestige ~ s(education,bs='cr') + s(income,bs='cr'),data = Prestige)
```

What `s()` does is basically building a "smooth" term for each of the
covariates I am putting in. The behavior is very similar to
`smooth.spline` we have seen last time, no need to set the number of
knots, nor (like in the `gam` package) the equivalent degrees of
freedom; `mgcv` will take care of everything for you. With `bs` we
indicate the (penalized) smoothing basis to use: `cr` provides a cubic
spline basis defined by a modest sized set of knots spread evenly
through the covariate values. They are penalized by the conventional
intergrated square second derivative cubic spline penalty. The fitting
is based on penalized likelihood, where the term to be penalized is the
second derivatives of the smooths.

Let us look at the models summary

```{r}
summary(model_gam)
```

If you are familiar with the output of a glm object, the one above
should not surprise you that much. The only slight exotic part may be
the "approximate significance of smooth terms": Effective degrees of
freedom (edf) is a summary statistic of GAM and it reflects the degree
of non-linearity of a curve. Notice that in this case the adjusted R
squared is better, even without the interaction term.

Let us look at the residuals:

```{r}
hist(model_gam$residuals)
qqnorm(model_gam$residuals)
```

Normality test:

```{r}
shapiro.test(model_gam$residuals)
```

Not bad at all! Notice that `mgcv` works with smoothing bases;
nevertheless, the beauty of GAMs is that we can use every type of
univariate smoother as building blocks for fitting an additive model. We
would for example like to employ natural cubic splines for providing
nonparametric components in our GAMs. In this case, we do not actually
need anything new then the old but gold `lm` function and the basis
matrix generator contained in the splines package.

```{r}
model_gam_ns <-
  lm(prestige ~ ns(education, df = 3) + ns(income, df = 3), data = Prestige)
```

In most situations, the differences in the GAMs obtained using smoothing
splines versus natural splines are small, let us look at the residuals
scatterplot for the two models:

```{r}
plot(model_gam_ns$residuals,model_gam$residuals)
cor(model_gam_ns$residuals,model_gam$residuals)
```

Pretty much the same fit!

GAMs make the effect interpretation of each regressor to the response
variable easy to understand. Again, it works very much like the analysis
of a linear model: I look at the contribution of a single predictor
holding all the others fixed. This is graphically best accomplished with
the plot method conveniently provided by the `mgcv` package.

```{r}
plot(model_gam)
```

The same can be achieved for the natural splines model, with a tiny
workaround (programming tip!)

```{r}
gam::plot.Gam(model_gam_ns, se=TRUE)
```

Again, we see that there are basically no difference between `model_gam`
and `model_gam_ns`.

We have noticed already at the very beginning that the relationship
between prestige and education seems fairly linear. We now test whether
a smooth function is needed for education or if we can be satisfied with
a linear contribution. The reduced model reads:

```{r}
model_gam_reduced=gam(prestige ~ education + s(income,bs='cr'),data = Prestige)
summary(model_gam_reduced)
```

The linear contribution of education is highly significant, but which
model is better? Let us perform a formal test as we would do if we
needed to perform nested model selection with lm: the anova function
serves the purpose.

```{r}
anova(model_gam_reduced,model_gam, test = "F")
```

Of course we could have had it hard-coded:

```{r, class.source="extracode"}
N <- nrow(Prestige)
RSS_reduced <- deviance(model_gam_reduced)
RSS_full <- deviance(model_gam)
df_full <- N-sum(summary(model_gam)$s.table[,2])-1 # -1 for the intercept
df_reduced <- N-sum(summary(model_gam_reduced)$s.table[,2])-2 # -2 for the intercept and the linear contribution of education
df_difference <- df_reduced-df_full
F_value_manual <- ((RSS_reduced-RSS_full)/df_difference)/(RSS_full/summary(model_gam)$residual.df)
pf(F_value_manual, df1 = df_difference,df2 = df_full,lower.tail = FALSE)
```

It seems that a nonlinear component for education is needed after all.
Important: if my residuals were not normal I would have needed to
perform a non-parametric test, but you now know how to do it right? More
on this later.

Performing prediction for GAMs is as simple as it always has been:

```{r, webgl=TRUE}
pred_gam=predict(model_gam,newdata=grid)

persp3d(education.grid,income.grid,pred_gam,col='grey30')
with(Prestige,points3d(education,income,prestige,col='black',size=5))
```

We can include interactions in a GAM as well: we could either employ a
simpler approach (the one we will subsequently use) by defining an
interaction term and smooth it, or we could employ more sophisticated
bidimensional splines, like thin plate spline ([Wood,
2003](https://rss.onlinelibrary.wiley.com/doi/10.1111/1467-9868.00374)).
In general, it is better to consider bivariate splines when the two
dimensions are tightly related (e.g., different coordinates in the
space).

```{r, webgl=TRUE}
model_gam_inter = gam(prestige ~ s(education, bs = 'cr') + 
                        s(income, bs ='cr') + 
                        s(I(income * education), bs = 'cr'),
                      data = Prestige)
# model_gam_inter=gam(prestige ~ s(education,bs='cr') + s(income,bs='cr')+ s(education,income,bs='tp'),data = Prestige) # thin plate spline
pred_inter = predict(model_gam_inter,
                     newdata = data.frame(grid, inter = grid$education * grid$income))

persp3d(education.grid, income.grid, pred_inter, col = 'grey30')
with(Prestige,
     points3d(education, income, prestige, col = 'black', size = 5))
```

## Exam like exercise 1 (2021/01/22)

Dr. Simoni, a data scientist, just had his second kid (a boy, 58 cm
tall) and he is interested to know how tall his child will be at 25
years of age, given his height at birth. To do so, he has collected the
height at birth of 100 males, alongside the height they reached when
they were 25. He has collected these data in the file *boyheight.rda*,
*height.25* is the height [cm] at 25 years of age, while *height.b* is
the length of the newborn [cm]. Assume that
$$height25 = f (height.b) + \varepsilon.$$

1.  Build a degree 1 regression spline model, with breaks at the 25th
    and 75th percentile of *height.b*, to predict the height at 25 from
    the height at birth. Provide a plot of the regression line, compute
    the pointwise prediction (round it at the second decimal digit) for
    the height at 25 of Dr. Simoni newborn child, and calculate, using a
    bootstrap approach on the residuals, the bias, variance and Mean
    Squared Error (MSE) of such prediction

2.  Dr. Simoni is not particularly satisfied with the predictions of
    such a simple model, and would like you to do more. So build a
    prediction model for the height at 25 based on a smoothing spline of
    order 4 (select the lambda parameter via Leave-One-Out CV). Report
    the optimal lambda value (2 decimal digits), provide a plot of the
    regression line, alongside the point-wise prediction of the height
    at 25 of Dr. Simoni's kid, and calculate using a bootstrap approach
    on the residuals the bias, variance and MSE of such prediction (fix
    the lambda value to the one obtained via Leave-One-Out CV).

### Solution

1.  

```{r}
load(here::here("Block III - Nonparametric regression/data/boyheight.rda"))
plot(x=height.b, y=height.25)

# Breaks at the 25th and 75th percentile of *height.b*
  
knots_heigth <- quantile(height.b,probs = c(.25,.75))

# Build a degree 1 regression spline model

fit_spline <- lm(height.25~bs(height.b,knots = knots_heigth,degree = 1))
new_data_seq <- seq(min(height.b),max(height.b),length.out=100)

# Predict the height at 25 from the height at birth
preds=predict(fit_spline, newdata = list(height.b=new_data_seq),se=T)
se.bands=cbind(preds$fit +2* preds$se.fit ,preds$fit -2* preds$se.fit)
```

```{r}
# Plot of the regression line
plot(y=height.25,x=height.b  ,cex =.5, col =" darkgrey " )
lines(new_data_seq,preds$fit ,lwd =2, col =" blue")
matlines(new_data_seq,se.bands ,lwd =1, col =" blue",lty =3)
```

```{r}
# Pointwise prediction
simoni_weigth <- 58
(pointwise_pred <- predict(fit_spline, newdata = list(height.b=simoni_weigth)))
```

```{r}
# Bias, variance and Mean Squared Error (MSE) of pointwise_pred
wrapper_f <- function(){
  height.25.boot=fitted+sample(residuals,n,replace=T)
  new_model=lm(height.25.boot ~ bs(height.b, knots=knots_heigth,degree=1))
  predict(new_model, newdata = list(height.b=simoni_weigth))
}

B <- 200
fitted=fit_spline$fitted.values
residuals=fit_spline$residuals
n <- length(height.25)
set.seed(100)

boot_d <- pbreplicate(n = B,expr = wrapper_f(),simplify = "vector")
```

```{r}
(variance_pred <- var(boot_d))
(bias_pred <- pointwise_pred-mean(boot_d))
(MSE_pred <- variance_pred +bias_pred^2)
```

2.  

```{r}
# smoothing spline of order 4 (cv=TRUE for LOOCV)
fit_smooth <- smooth.spline(x = height.b, y=height.25,cv = TRUE)

# optimal lambda value
round(fit_smooth$lambda,2)

# plot of the regression line
plot(y=height.25,x=height.b  ,cex =.5, col =" darkgrey " )
lines(fit_smooth$x, fit_smooth$y, col="blue", lwd=2)

fitted_smooth <- predict(fit_smooth, height.b)$y
fitted_res <- height.25-fitted_smooth

pred_smooth_simoni <- predict(fit_smooth, simoni_weigth)
# pointwise prediction of the height at 25 of Dr. Simoni’s kid
pred_smooth_simoni$y
```

```{r}
boot_smooth=numeric(B)
set.seed(100)
for(sample in 1:B){
  
  height.25.boot=fitted_smooth+sample(fitted_res,n,replace=T)
  new_model=smooth.spline(y = height.25.boot,x =  height.b, lambda = fit_smooth$lambda)
  boot_smooth[sample]=predict(new_model, simoni_weigth)$y
}


(variance_pred_smooth <- var(boot_smooth))
(bias_pred_smooth <- pred_smooth_simoni$y-mean(boot_smooth))
(MSE_pred_smooth <- variance_pred_smooth +bias_pred_smooth^2)
```

## Exam like exercise 2 (2021/06/14)

Dr. Bisacciny, Ph.D is becoming increasingly worried about the fact that
the bees that he decided to place near a corn field close to the Italian
city of Milan tend to die way earlier than the ones that are placed
close to the small village of Jovençan, in Aosta Valley. He suspects
that additional factors that influence the survival time of a the bee
are its productivity, and its weight. For this reason, he decides to run
an experiment, in which he selects 10,000 bees, placing 5,000 of them in
Aosta Valley and 5,000 in Milan. Dr. Bisacciny runs the experiment for
50 days, during which he annotates when a bee passes away. After the end
of the experiment, he starts to analyse the data. In the file ex02.rda
you can find a dataframe with information about the status of the bee (1
if alive, 2 if dead), the survival time of the bee (surv.time), its
weight (weight) and productivity (prod) as well as its being in Jovençan
or in Milan. Modeling this data as i.i.d. realizations of a four
dimensional random variable:

1.  Build an additive model for log(surv.time), where log() is the
    natural logarithm, using cubic b-spline terms for the main effects,
    a dummy term for location and no interactions. After having written
    in proper mathematical terms the additive model you have estimated,
    report the adjusted $R^2$ and the p-values of the tests. Comment the
    result (assume the residuals to be normal).
2.  After having reduced the model to its significant terms and use this
    model to provide a pointwise prediction for the average log-survival
    time for a bee that lives in Milan or in Jovençan. Is it necessary
    to use the gam machinery to estimate the reduced model?

### Solution

```{r}
load(here::here("Block III - Nonparametric regression/data/ex02.rda"))
head(bee)
```

```{r}
fit_gam <- gam(log(surv.time)~s(weight,bs = "cr")+s(prod,bs = "cr")+location,data = bee)
```

Estimated model

```{=tex}
\begin{equation}
log(surv.time) =\beta_{0}+f\left(weight\right)+f\left(prod\right)+\beta_1 location_{Milan}+\epsilon
\end{equation}
```
```{r}
table_fit_gam <- summary(fit_gam)
# adjusted R2
(r_2_squared <- table_fit_gam$r.sq)

# p-values of the tests (parametric coefficients)
table_fit_gam$p.pv

# p-values of the tests (smooth terms)
table_fit_gam$s.pv
```

```{r}
plot(fit_gam)
```

```{r}
hist(fit_gam$residuals)
```

2.  

```{r}
# Remove weight
fit_gam_reduced_1 <- gam(log(surv.time)~s(prod,bs = "cr")+location,data = bee)
anova(fit_gam_reduced_1,fit_gam, test = "F")

# Remove prod
fit_gam_reduced_2 <- gam(log(surv.time)~location,data = bee)
anova(fit_gam_reduced_2, fit_gam_reduced_1, test = "F")
```

I can use a simple linear model with just a term: location. No need to
use the gam machinery anymore.

```{r}
fit_lm <- lm(log(surv.time)~location,data = bee)

# Avg log surv time in Milan
sum(fit_lm$coefficients)
# mean(log(bee$surv.time[bee$location=="Milan"])) # alternatively

# Avg log surv time in Jovencan
fit_lm$coefficients[1]
# mean(log(bee$surv.time[bee$location=="Jovencan"])) # alternatively
```

More concise way using `tapply` and functional programming

```{r}
tapply(log(bee$surv.time), bee$location, mean)
```
